% figures in chp.tex files need to have paths relative to main.tex
\documentclass[a4paper]{report}

\usepackage{a4}
\usepackage{epsfig}
\usepackage{graphics,graphicx}
\usepackage[lmargin=4cm,rmargin=2cm,tmargin=3.5cm,bmargin=3.5cm]{geometry}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{tabulary}
\usepackage{txfonts}
\usepackage{mathtools}
\usepackage{upgreek}
\usepackage{subfig}
\usepackage{float}
\usepackage{longtable}
  \newtheorem{lem}{Lemma}
  \newtheorem{thm}{Theorem}
  
  \newcounter{appendix}[chapter]

\makeatletter

\newcommand\bappendix{%
  \addtocontents{toc}{\protect\addvspace{10pt}Appendices}
  \begingroup
  \renewcommand\section{\stepcounter{appendix}%
    \renewcommand\thesection{\thechapter.\Alph{appendix}}
    \@startsection {section}{1}{\z@}%
      {-3.5ex \@plus -1ex \@minus -.2ex}%
      {2.3ex \@plus.2ex}%
      {\normalfont\Large\bfseries}}}
\makeatother
\newcommand\eappendix{\endgroup}

%\usepackage{lineno}
%\setcounter{page}{1}

%\pagenumbering{arabic}

\usepackage{fancyhdr}

\newcommand{\rr}{\raggedright}
\newcommand{\tn}{\tabularnewline}
\newtheorem{mydef}{Definition}

\begin{document}

\pagestyle{fancy}
\renewcommand{\chaptermark}[1]{%
\markboth{\MakeUppercase{%
\chaptername\ \thechapter.%
\ #1}}{}}
\fancyhead{}
%\fancyfoot{}
\fancyhead[R]{\leftmark}

%\setpagewiselinenumbers

\input{titlepage.tex}

\newpage

\vspace*{\fill}

\begin{centering}

\textit{I, Stuart Nattrass, confirm that the work presented in this thesis is my own. Where information has been derived from other sources, I confirm that this has been indicated in the thesis.}

\end{centering}

\vspace*{\fill}


\newpage

%\newpage


%\vspace*{\fill}

%\begin{verse}
%\textit{``Confusion is sex''}\end{verse}\begin{flushright}
%Sonic Youth
%\end{flushright}

%\vspace*{\fill}

\newpage

%\pagenumbering{arabic}

\doublespacing

\chapter*{Acknowledgements}

Thanks to my supervisors, David Murrell and Stephen Baigent, for their guidance, ideas, and inspiration. They have made this entire process an enjoyable one. I have learnt a great deal from both of you, and I am incredibly grateful for everything.

To those who have shared an office, or drinks, with me, thank you for teaching a mathematician biology, and keeping me entertained along the way. Special mentions to Lawrence Bellamy and Aaron Towlson, for welcoming me into the department with pints and open arms, Fraser Murdoch and Claire Asher for keeping the nights out silly, Anton Fl\"{u}gge, Kate Brown, Pascale Gerbault, and latterly Henry Ferguson-Gow, Tim Lucas and the rest of the Kate Jones group, with whom I shared an office all too briefly. Sam Tazzyman's help from afar during the writing process has also been invaluable. And Jane Dempster, lovely Wo-Ho mum.

Thanks also to my non-science friends, for reminding me that there is a whole world out there.

Finally thanks to Mam and Dad, for putting up with my refusal to stop being a student every step of the way. 

\chapter*{Abstract}
This thesis uses theoretical models to answer questions that evolve around the ecological processes that maintain diversity. Chapter 1 studies the importance of competitive asymmetry on the likelihood of coexistence. Considering a competition-fecundity trade-off, we find conditions that allow coexistence of two species, and show the likelihood of these species coexisting increases with competitive asymmetry. Further, we demonstrate that asymmetry allows for the coexistence of an arbitrarily large number of species, although the likelihood of coexistence decreases as more species are considered. Chapters 2, 3 and 4 are all concerned with the effects of disturbance on diversity. In Chapter 2, we demonstrate that the frequency and intensity of disturbance events can affect the community dynamics in different ways. We demonstrate that the diversity-disturbance relationship can take several different structures as frequency and intensity interact to determine overall disturbance regimes. Chapter 3 examines which of three predicted trade-offs are likely to contribute most to coexistence. In particular, we derive the likelihood of two species coexisting as defence against disturbance; fecundity and juvenile growth rates vary between species. We show that a fecundity-defence trade-off is particularly sensitive to parameters, and is not expected to contribute to diversity in large communities. Further, we conclude that a fecundity-growth trade-off contributes more to the maintenance of diversity than a growth-defence trade-off, although the latter Is expected to have some effect. In Chapter 4, we consider the effects of disturbance extent on the likelihood of three species coexistence. We demonstrate that in a single, well mixed community a fecundity-growth trade-off cannot support three species. We then use a metapopulation structure to show that spatial heterogeneity in disturbance regimes can increase diversity; showing that a community protected from disturbance can contribute to increase regional diversity, and increase diversity in a neighbouring patch with disturbance.

\tableofcontents
\listoffigures
\listoftables

%\pagenumbering{arabic}

%\linenumbers
    \newpage
    \setcounter{chapter}{-1}
\chapter{Introduction}
\input{Introduction}

\newpage
\chapter{Quantifying the probability of coexistence in communities with asymmetric competition}
\chaptermark{Quantifying the probability of coexistence}
This work was conducted in collaboration with David J Murrell and Stephen Baigent, and was previously published in Bulletin of Mathematical Biology \citep{nattrass2012quantifying}
\input{C1Asymmetry}

\newpage
\chapter{The effect of disturbance intensity and frequency on coexistence via a fecundity-growth trade-off}
\chaptermark{The effect of disturbance on coexistence}
This work was conducted in collaboration with David J Murrell and Stephen Baigent
\input{C2Disturbance}

\newpage
\chapter{Which trade-off contributes most to biodiversity? A comparison of trade-off models}
\chaptermark{Which trade-off contributes most to biodiversity?}
This work was conducted in collaboration with David J Murrell and Stephen Baigent
\input{C3Comparison}

\newpage
\chapter{Multi-species coexistence: the effects of disturbance refugia on species richness}
\chaptermark{Multi-species coexistence}
This work was conducted in collaboration with David J Murrell and Stephen Baigent
\input{C4Patch}


\newpage
\chapter{General Conclusion}
\input{Conc}

   \newpage
\bibliographystyle{spbasic}
\bibliography{Papers.bib}

\pagebreak


\end{document}
